A tutorial on spatiotemporal partially observed Markov process models via the R package spatPomp

### 2024-04-18 ms.Rnw
arxiv replacement

\documentclass[article,pdftex,shortnames,nojss]{jss}

\usepackage{amsmath,amssymb}
\usepackage{bbm} % for \mathbbm{1}
\graphicspath{{figure/}}

@@

@@

\author{Kidus Asfaw\\Univ. of Michigan \And
        Joonha Park\\Univ. of Kansas \And
        Aaron A. King\\Univ. of Michigan \And
        Edward L. Ionides\\Univ. of Michigan}
\Plainauthor{Kidus Asfaw, Joonha Park, Aaron A. King and Edward L. Ionides}

\title{Partially observed Markov processes with spatial structure via the \proglang{R} package \pkg{spatPomp}\\}
\Plaintitle{Spatially Coupled Partially Observed Markov Processes via the R Package spatPomp}
\Shorttitle{\pkg{{\spatPomp}}: Spatially Coupled Partially Observed Markov Processes in \proglang{R}}

\Abstract{
We introduce a computational framework for modeling and statistical inference on high-dimensional dynamic systems.
Our primary motivation is the investigation of metapopulation dynamics arising from a collection of spatially distributed, interacting biological populations.
To make progress on this goal, we embed it in a more general problem: inference for a collection of interacting partially observed nonlinear non-Gaussian stochastic processes.
Each process in the collection is called a unit;
in the case of spatiotemporal models, the units correspond to distinct spatial locations.
The dynamic state for each unit may be discrete or continuous, scalar or vector valued.
In metapopulation applications, the state can represent a structured population or the abundances of a collection of species at a single location.
We consider models where the collection of states has a Markov property.
A sequence of noisy measurements is made on each unit, resulting in a collection of time series.
A model of this form is called a spatiotemporal partially observed Markov process (SpatPOMP).
The \proglang{R} package \pkg{\spatPomp} provides an environment for implementing SpatPOMP models, analyzing data using existing methods, and developing new inference approaches.
Our presentation of \pkg{\spatPomp} reviews various methodologies in a unifying notational framework.
We demonstrate the package on a simple Gaussian system and on a nontrivial epidemiological model for measles transmission within and between cities.
We show how to construct user-specified SpatPOMP models within \pkg{\spatPomp}.

\noindent
This version was compiled on {\today} using \pkg{spatPomp} \Sexpr{packageVersion("spatPomp")} with \pkg{pomp} \Sexpr{packageVersion("pomp")} and \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)} from source code at \code{https://github.com/ionides/spatPomp-article}.
It is provided under the Creative Commons Attribution License. 
}

\Keywords{Markov processes, hidden Markov model, state space model, stochastic dynamical system, maximum likelihood, plug-and-play, spatiotemporal data, mechanistic model, sequential Monte Carlo, \proglang{R}}
\Plainkeywords{Markov processes, hidden Markov model, state space model, stochastic dynamical system, maximum likelihood, plug-and-play, spatiotemporal data, mechanistic model, sequential Monte Carlo, R}

\Address{
  Kidus Asfaw, Edward L. Ionides\\
  University of Michigan\\
  48109 Michigan, United States of America\\
  E-mail: \email{kasfaw@umich.edu}, \email{ionides@umich.edu}\\
  URL: \url{https://www.stat.lsa.umich.edu/~ionides/}\\

  Joonha Park\\
  Department of Mathematics\\
  University of Kansas\\
  66045 Kansas, United States of America\\
  E-mail: \email{j.park@ku.edu}\\
  URL: \url{https://people.ku.edu/~j139p002}\\

  Aaron A. King\\
  Department of Ecology \& Evolutionary Biology\\
  Center for the Study of Complex Systems\\
  University of Michigan\\
  48109 Michigan, United States of America\\
  E-mail: \email{kingaa@umich.edu}\\
  URL: \url{https://kinglab.eeb.lsa.umich.edu/}\\

@@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    comment="",
    error=TRUE,
    echo=TRUE,
    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

@@

@@

A spatiotemporal partially observed Markov process (SpatPOMP) model consists of incomplete and noisy measurements of a latent Markov process having spatial as well as temporal structure.
A SpatPOMP model is a special case of a vector-valued partially observed Markov process (POMP) where the latent states and the measurements are indexed by a collection of spatial locations known as units.
Many biological, social and physical systems have the spatiotemporal structure, dynamic stochasticity and imperfect observability that characterize SpatPOMP models.
This paper describes the \pkg{spatPomp} package \citep{R:spatPomp}, available for \proglang{R} \citep{R} from the Comprehensive R Archive Network (CRAN).
The objective of the \pkg{\spatPomp} package is to facilitate model development and data analysis in the context of the general class of SpatPOMP models, enabling scientists to separate the scientific task of model development from the statistical task of providing inference tools.

@@

A SpatPOMP model is characterized by the transition density for the latent Markov process and unit-specific measurement densities.
Once these elements are specified, calculating and simulating from all joint and conditional densities are well defined operations.
However, different statistical methods vary in the operations they require.
Some methods require only simulation from the transition density whereas others require evaluation of this density.
Some methods avoid working with the model directly, replacing it by an approximation, such as a linearization.
For a given model, some operations may be considerably easier to implement and so it is useful to classify inference methods according to the operations on which they depend.
In particular, an algorithm is said to be \emph{plug-and-play} if it utilizes simulation of the latent process but not evaluation of transition densities \citep{breto09,he10}.
Simulators are relatively easy to implement for many SpatPOMP models, and so plug-and-play methodology facilitates the investigation of a variety of models that may be scientifically interesting but mathematically inconvenient.
Modern plug-and-play algorithms can provide statistically efficient likelihood-based or Bayesian inference.
The computational cost of plug-and-play methods may be considerable, due to the large number of simulations involved.
Nevertheless, the practical utility of plug-and-play methods for POMP models has been amply demonstrated in scientific applications.
In particular, plug-and-play methods implemented using \pkg{pomp} have facilitated various scientific investigations \citep[e.g.,][]{king08,bhadra11,shrestha11,shrestha13,earn12,roy13,blackwood13,blackwood13b,he13,breto14,blake14,martinez-bakker15,bakker16,becker16,buhnerkempe17,ranjeva17,marino19,pons-salort18,becker19,kain20,stocks20}.
The \pkg{spatPomp} package has been used to develop and demonstrate plug-and-play methodology for SpatPOMP models \citep{ionides21,ionides22} and scientific applications of these methods are anticipated.

The remainder of this paper is organized as follows.
\Cref{sec:background} defines mathematical notation for SpatPOMP models and relates this to their representation as objects of \class{\spatPomp} in the \pkg{\spatPomp} package.
\Cref{sec:filtering} introduces likelihood evaluation via several spatiotemporal filtering methods.
\Cref{sec:inference} describes parameter estimation algorithms which build upon these filtering techniques.
\Cref{sec:brownian} constructs a simple linear Gaussian SpatPOMP model and uses this example to illustrate statistical inference.
\Cref{sec:measles} presents the construction of spatially structured compartment models for population dynamics, in the context of coupled measles dynamics in UK cities;
this demonstrates the kind of nonlinear stochastic system primarily motivating the development of \pkg{\spatPomp}.
Finally, \cref{sec:conclusion} discusses extensions and applications of \pkg{\spatPomp}.

@@

      log likelihood component,

@@

    log likelihood,

@@

\begin{algorithm}[tb]
  \caption{\code{enkf(P,Np{\argequals}$J$)}, using notation from \cref{tab:notation} where \code{P} is a `\code{\spatPomp}' object equipped with
    \code{rprocess},
    \code{eunit\_measure},
    \code{vunit\_measure},
    \code{rinit},
    \code{coef},
    \code{obs}.
    \label{alg:enkf}
  }
  \KwIn{
    simulator for $f_{\vec{X}_{\time}|\vec{X}_{\time-1}}(\vec{x}_{\time} \given \vec{x}_{\time-1}\giventh\vectheta)$ and $f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)$;
    evaluator for $\emeasure_{\unit}(X_{\unit, \time},\vectheta)$ and $\vmeasure_{\unit}(X_{\unit,\time},\vectheta)$;
    parameter, $\vectheta$;
    data, $\data{\vec{y}}_{1:\Time}$;
    number of particles, $J$.
  }
  initialize filter particles,
  $\vec{X}_{0}^{F,\np}\sim {f}_{\vec{X}_{0}}\left(\mydot\giventh{\vectheta}\right)$
  for $\np$ in $\seq{1}{\Np}$
  \;
      \For{$\time\ \mathrm{in} \ \seq{1}{\Time}$}{
        prediction ensemble,
    $\vec{X}_{\time}^{P,\np}\sim {f}_{\vec{X}_{\time}|\vec{X}_{\time-1}}\big(\mydot|\vec{X}_{\time-1}^{F,\np};\vectheta\big)$
    for $\np$ in $\seq{1}{\Np}$
    \nllabel{alg:enkf:prediction}
    \;
        centered prediction ensemble, $\tilde{\vec{X}}_{\time}^{P,\np} =
        \vec{X}_{\time}^{P,\np} - \frac{1}{\Np}\sum_{\altNp=1}^{\Np}\vec{X}_{\time}^{P,\altNp}$
       for $\np$ in $\seq{1}{\Np}$
    \;
        forecast ensemble, $\vec{\hat{Y}}^{\np}_{\!\time}=\emeasure_{\unit}(X_{\unit,\time}^{P,\np},\vectheta)$
       for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:forecast}
        \;
	forecast mean, $\overline{\vec{Y}}_{\!\time}=\frac{1}{\Np}\sum_{\np=1}^{\Np}\vec{\hat{Y}}^{\np}_{\!\time}$
	\;
        centered forecast ensemble, $\vec{\tilde{Y}}^{\np}_{\time} =
        \vec{\hat{Y}}^{\np}_{\!\time} - \overline{\vec{Y}}_{\!\time}$
       for $\np$ in $\seq{1}{\Np}$
        \;
        forecast measurement variance,
	$R_{\unit,\altUnit} = \mathbbm{1}_{\unit,\altUnit} \,
	  \frac{1}{\Np}\sum_{\np=1}^{\Np}
	    \vmeasure_{\unit}\big(
	      \vec{X}_{\unit,\time}^{P,\np},\vectheta\big)$
         \nllabel{alg:enkf:cond:var}
	 for $\unit, \altUnit$ in $\seq{1}{\Unit}$
        \;
        forecast estimated covariance, $\Sigma_{Y}= \frac{1}{\Np-1}\sum_{\np=1}^{\Np}(\vec{\tilde{Y}}^{\np}_{\!\time})(\vec{\tilde{Y}}^{\np}_{\!\time})^T + R$
        \;
        prediction and forecast sample covariance, $\Sigma_{XY}=\frac{1}{\Np-1}\sum_{\np=1}^{\Np}(\tilde{\vec{X}}_{\time}^{P,\np})(\vec{\tilde{Y}}^{\np}_{\!\time})^T$
        \;
        Kalman gain, $K = \Sigma_{XY}\Sigma_{Y}^{-1}$
        \;
        artificial measurement noise, $\vec{\artNoise}_{\time}^{\np}\sim \normal(\vec{0},R)$
	for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:artificial:noise}
        \;
        errors, $\vec{r}_{\time}^{\np}= \vec{\hat{Y}}^{\np}_{\!\time} - \data{\vec{y}}_{\time}$
	for $\np$ in $\seq{1}{\Np}$
        \;
        filter update,
        $\vec{X}_{\time}^{F,\np} = \vec{X}_{\time}^{P,\np} +
              K\big( \vec{r}_{\time}^{\np}+\vec{\artNoise}_{\time}^{\np}\big)$
	for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:update}
	\;
	$\loglik_{\time}=\log \big[ \phi\big(\data{\vec{y}}_{\time} \giventh \overline{\vec{Y}}_{\!\time} , \Sigma_{Y} \big) \big]$ where $\phi(\cdot\giventh \vec{\mu},\Sigma)$ is the $\normal(\vec{\mu},\Sigma)$ density.
      }
  \KwOut{
    filter sample, $\vec{X}^{F,1:\Np}_{\time}$, for $n$ in $\seq{1}{N}$;
    log likelihood estimate, $\loglik^{\mbox{\tiny{EnKF}}}=\sum_{\time=1}^{\Time} \loglik_{\time}$
  }
  \KwCplx{$\bigO(\Np\Unit\Time)$}
\end{algorithm}

@@

\begin{algorithm}[t]
 \caption{\code{bpfilter(P,Np{\argequals}$\Np\!$,block\_list{\argequals}$\blocklist$)}
    using notation from \cref{tab:notation} where \code{P} is a
     `\code{\spatPomp}' object equipped with
     \code{rprocess},
     \code{dunit\_measure},
     \code{rinit},
     \code{obs},
     \code{coef}.
   }\label{alg:bpfilter}
  \KwIn{
    simulator for $f_{\vec{X}_{\time}|\vec{X}_{\time-1}}(\vec{x}_{\time}\given \vec{x}_{\time-1}\giventh\vectheta)$ and $f_{\vec{X}_0}(\vec{x}_{0}\giventh\vectheta)$;
   number of particles, $\Np$;
    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time}\giventh\vectheta)$;
    data, $\data{\vec{y}}_{1:\Time}$;
    parameter, $\vectheta$;
     blocks, $\blocklist_{1:\Block}$;
  }
initialization, $\vec{X}_{0}^{F,\np}\sim {f}_{\vec{X}_{0}}\left(\mydot\giventh{\vectheta}\right)$
for $\np$ in $\seq{1}{\Np}$
\;
\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{ % For each time step
  prediction,
  $\vec{X}_{\time}^{P,\np}\sim {f}_{\vec{X}_{\time}|\vec{X}_{\time-1}}\big(\mydot|\vec{X}_{\time-1}^{F,\np};{\vectheta}\big)$
  for $\np$ in $\seq{1}{\Np}$
    \;
    block weights,
          $\displaystyle \blockweight_{\block,\time}^{\np}=
            \prod_{\unit \in \blocklist_{\block}}
            f_{Y_{\unit,\time}|X_{\unit,\time}}
            \big(
              \data{y}_{\unit,\time}\given X^{P,\np}_{\unit,\time} \giventh \vectheta
            \big)$
    for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
          \nllabel{alg:bpfilter:blockweights}\;
   resampling indices,
      $\resampleIndex^{\np}_{\block,\time}$ with
      $\prob\left[\resampleIndex^{\np}_{\block,\time}=i \right] =
      {\blockweight}^{i}_{\block,\time}
      \Big/\sum_{q=1}^{\Np}\blockweight^{q}_{\block,\time}$
      for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
    \;
    resample,
      $X_{\blocklist_{\block},\time}^{F,\np}=
      X_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}}$
      for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
      \nllabel{alg:bpfilter:resample}
    \;
}%End timestep for Loop
\KwOut{
    log likelihood,
      $\loglik^{\mbox{\tiny{BPF}}}(\vectheta)= \sum_{\time=1}^{\Time}
        \sum_{\block=1}^{\Block}
\log\Big(
        \frac{1}{\Np}
        \sum_{\np=1}^{\Np}
          \blockweight^{\np}_{\block,\time}
      \Big)$, filter particles $\vec{X}^{F,1:\Np}_{1:\Time}$
}
\KwCplx{$\bigO(\Np \Unit \Time)$}
\end{algorithm}

@@

The adapted bagged filter \citep{ionides21} combines many independent particle filters.
This is called \emph{bagging}, (\emph{b}ootstrap \emph{agg}regat\emph{ing}), since a basic particle filter is also called a bootstrap filter.
The adapted distribution is the conditional distribution of the latent process given its current value and the subsequent observation \citep{johansen08}.
In the adapted bagged filter, each bootstrap replicate makes a Monte Carlo approximation to a draw from the adapted distribution.
Thus, in the pseudocode of \cref{alg:abf}, $\vec{X}^{\IF,\rep}_{0:\Time}$ is a Monte Carlo sample targeting the adapted sampling distribution,
\begin{equation}
  \label{adapteddistribution}
f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)\prod_{\time=1}^\Time
f_{\vec{X}_{\time}|\vec{Y}_{\time},\vec{X}_{\time-1}}
(\vec{x}_{\time}\given \data{\vec{y}}_{\time},\vec{x}_{\time-1} \param\vectheta).
\end{equation}
Each adapted simulation replicate is constructed by importance sampling using proposal particles $\{\vec{X}^{\IP,\rep,\np}_{\time}\}$.
The ensemble of adapted simulation replicates are then weighted using data in a spatiotemporal neighborhood of each observation to obtain a locally combined Monte Carlo sample targeting the filter distribution, with some approximation error due to the finite spatiotemporal neighborhood used.
This local aggregation of the bootstrap replicates also provides an evaluation of the likelihood function.

@@

ABF can be combined with the guided intermediate resampling technique used by GIRF to give an algorithm called ABF-IR \citep{ionides21} implemented as \code{abfir}.

@@

  parameter log likelihoods, 

@@

The \code{measles()} and \code{he10()} models are similar, but differ in details.
For \code{measles()}, the data consist of biweekly counts and parameters are all shared between units, matching the analysis of \citet{park20} and \citet{ionides21}.
For \code{he10()}, data are weekly and parameters can be shared or unit-specific, matching the analysis of \citet{he10} and \citet{ionides22}.
We write the model for the general case where all parameters are unit-specific, noting that it is a relevant data analysis question to determine when parameter dependence on $\unit$ can be omitted.

>>>>

\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{bbm} % for \mathbbm{1}
\graphicspath{{figure/}}

\usepackage{natbib}
\bibliographystyle{apalike}
%\usepackage{url}

@@

\newcommand\proglang[1]{\textsf{#1}}
\newcommand\code[1]{\texttt{#1}}
\newcommand\pkg[1]{\texttt{#1}}
\newcommand\E{\mathsf{E}}
\newcommand\Prob{\mathsf{P}}

@@

\author{Kidus Asfaw$^1$, Joonha Park$^2$, Aaron A. King$^1$ and Edward L. Ionides$^{1*}$\\
\vspace{2mm}
\small{$^1$University of Michigan, $^2$University of Kansas, $^*$correspondence to \texttt{ionides@umich.edu} } }

\title{A tutorial on spatiotemporal partially observed Markov process models via the R package spatPomp}

\date{}

@@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    comment="",
    error=TRUE,
    echo=TRUE,
#    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

@@

\maketitle

\begin{abstract}
We describe a computational framework for modeling and statistical inference on high-dimensional stochastic dynamic systems.
Our primary motivation is the investigation of metapopulation dynamics arising from a collection of spatially distributed, interacting biological populations.
To make progress on this goal, we embed it in a more general problem: inference for a collection of interacting partially observed nonlinear non-Gaussian stochastic processes.
Each process in the collection is called a unit;
in the case of spatiotemporal models, the units correspond to distinct spatial locations.
The dynamic state for each unit may be discrete or continuous, scalar or vector valued.
In metapopulation applications, the state can represent a structured population or the abundances of a collection of species at a single location.
We consider models where the collection of states has a Markov property.
A sequence of noisy measurements is made on each unit, resulting in a collection of time series.
A model of this form is called a spatiotemporal partially observed Markov process (SpatPOMP).
The \proglang{R} package \pkg{\spatPomp} provides an environment for implementing SpatPOMP models, analyzing data using existing methods, and developing new inference approaches.
Our presentation of \pkg{\spatPomp} reviews various methodologies in a unifying notational framework.
We demonstrate the package on a simple Gaussian system and on a nontrivial epidemiological model for measles transmission within and between cities.
We show how to construct user-specified SpatPOMP models within \pkg{\spatPomp}.

\noindent
This version was compiled on {\today} using \pkg{spatPomp} \Sexpr{packageVersion("spatPomp")} with \pkg{pomp} \Sexpr{packageVersion("pomp")} and \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}. Source code for this article is at \code{https://github.com/ionides/spatPomp-article}.
Materials are provided under the Creative Commons Attribution License.

\end{abstract}

@@

A spatiotemporal partially observed Markov process (SpatPOMP) model consists of incomplete and noisy measurements of a latent Markov process having spatial as well as temporal structure.
A SpatPOMP model is a special case of a vector-valued partially observed Markov process (POMP) where the latent states and the measurements are indexed by a collection of spatial locations known as units.
Many biological, social and physical systems have the spatiotemporal structure, dynamic stochasticity and imperfect observability that characterize SpatPOMP models.
This paper discusses investigation of SpatPOMP models using the \pkg{spatPomp} software package \citep{R:spatPomp}, written in \proglang{R} \citep{R}.

@@

A SpatPOMP model is characterized by the transition density for the latent Markov process and unit-specific measurement densities.
Once these elements are specified, calculating and simulating from all joint and conditional densities are well defined operations.
However, different statistical methods vary in the operations they require.
Some methods require only simulation from the transition density whereas others require evaluation of this density.
Some methods avoid working with the model directly, replacing it by an approximation, such as a linearization.
For a given model, some operations may be considerably easier to implement and so it is useful to classify inference methods according to the operations on which they depend.
In particular, an algorithm is said to be \emph{plug-and-play} if it utilizes simulation of the latent process but not evaluation of transition densities \citep{breto09,he10}.
Simulators are relatively easy to implement for many SpatPOMP models, and so plug-and-play methodology facilitates the investigation of a variety of models that may be scientifically interesting but mathematically inconvenient.
Modern plug-and-play algorithms can provide statistically efficient likelihood-based or Bayesian inference.
The computational cost of plug-and-play methods may be considerable, due to the large number of simulations involved.
Nevertheless, the practical utility of plug-and-play methods for POMP models has been amply demonstrated in scientific applications.
In particular, plug-and-play methods implemented using \pkg{pomp} have facilitated various scientific investigations \citep[e.g.,][]{king08,bhadra11,shrestha11,shrestha13,earn12,roy13,blackwood13,blackwood13b,he13,breto14,blake14,martinez-bakker15,bakker16,becker16,buhnerkempe17,ranjeva17,marino19,pons-salort18,becker19,kain20,stocks20}.
The \pkg{spatPomp} package has been used to develop and demonstrate plug-and-play methodology for SpatPOMP models \citep{ionides22,ionides23,ning23-ibpf}.
Scientific applications are starting to emerge \citep{zhang22,wheeler24,li24}.

The remainder of this paper is organized as follows.
\Cref{sec:background} defines mathematical notation for SpatPOMP models and relates this to their representation as objects of \class{\spatPomp} in the \pkg{\spatPomp} package.
\Cref{sec:filtering} introduces likelihood evaluation via several spatiotemporal filtering methods.
\Cref{sec:inference} describes parameter estimation algorithms which build upon these filtering techniques.
\Cref{sec:brownian} constructs a simple linear Gaussian SpatPOMP model and uses this example to illustrate statistical inference.
\Cref{sec:measles} presents the construction of spatially structured compartment models for population dynamics, in the context of coupled measles dynamics in UK cities;
this demonstrates the kind of nonlinear stochastic system primarily motivating the development of \pkg{\spatPomp}.
\cref{sec:conclusion} is a concluding discussion.

@@

      log-likelihood component,

@@

    log-likelihood,

@@

\begin{algorithm}[tb]
  \caption{\code{enkf(P,Np{\argequals}$J$)}, using notation from \cref{tab:notation} where \code{P} is a `\code{\spatPomp}' object equipped with
    \code{rprocess},
    \code{eunit\_measure},
    \code{vunit\_measure},
    \code{rinit},
    \code{coef},
    \code{obs}.
    \label{alg:enkf}
  }
  \KwIn{
    simulator for $f_{\vec{X}_{\time}|\vec{X}_{\time-1}}(\vec{x}_{\time} \given \vec{x}_{\time-1}\giventh\vectheta)$ and $f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)$;
    evaluator for $\emeasure_{\unit}(X_{\unit, \time},\vectheta)$ and $\vmeasure_{\unit}(X_{\unit,\time},\vectheta)$;
    parameter, $\vectheta$;
    data, $\data{\vec{y}}_{1:\Time}$;
    number of particles, $J$.
  }
  initialize filter particles,
  $\vec{X}_{0}^{F,\np}\sim {f}_{\vec{X}_{0}}\left(\mydot\giventh{\vectheta}\right)$
  for $\np$ in $\seq{1}{\Np}$
  \;
      \For{$\time\ \mathrm{in} \ \seq{1}{\Time}$}{
        prediction ensemble,
    $\vec{X}_{\time}^{P,\np}\sim {f}_{\vec{X}_{\time}|\vec{X}_{\time-1}}\big(\mydot|\vec{X}_{\time-1}^{F,\np};\vectheta\big)$
    for $\np$ in $\seq{1}{\Np}$
    \nllabel{alg:enkf:prediction}
    \;
        centered prediction ensemble, $\tilde{\vec{X}}_{\time}^{P,\np} =
        \vec{X}_{\time}^{P,\np} - \frac{1}{\Np}\sum_{\altNp=1}^{\Np}\vec{X}_{\time}^{P,\altNp}$
       for $\np$ in $\seq{1}{\Np}$
    \;
        forecast ensemble, $\vec{\hat{Y}}^{\np}_{\!\time}=\emeasure_{\unit}(X_{\unit,\time}^{P,\np},\vectheta)$
       for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:forecast}
        \;
	forecast mean, $\overline{\vec{Y}}_{\!\time}=\frac{1}{\Np}\sum_{\np=1}^{\Np}\vec{\hat{Y}}^{\np}_{\!\time}$
	\;
        centered forecast ensemble, $\vec{\tilde{Y}}^{\np}_{\time} =
        \vec{\hat{Y}}^{\np}_{\!\time} - \overline{\vec{Y}}_{\!\time}$
       for $\np$ in $\seq{1}{\Np}$
        \;
        forecast measurement variance,
	$R_{\unit,\altUnit} = \mathbbm{1}_{\unit,\altUnit} \,
	  \frac{1}{\Np}\sum_{\np=1}^{\Np}
	    \vmeasure_{\unit}\big(
	      \vec{X}_{\unit,\time}^{P,\np},\vectheta\big)$
         \nllabel{alg:enkf:cond:var}
	 for $\unit, \altUnit$ in $\seq{1}{\Unit}$
        \;
        forecast estimated covariance, $\Sigma_{Y}= \frac{1}{\Np-1}\sum_{\np=1}^{\Np}(\vec{\tilde{Y}}^{\np}_{\!\time})(\vec{\tilde{Y}}^{\np}_{\!\time})^T + R$
        \;
        prediction and forecast sample covariance, $\Sigma_{XY}=\frac{1}{\Np-1}\sum_{\np=1}^{\Np}(\tilde{\vec{X}}_{\time}^{P,\np})(\vec{\tilde{Y}}^{\np}_{\!\time})^T$
        \;
        Kalman gain, $K = \Sigma_{XY}\Sigma_{Y}^{-1}$
        \;
        artificial measurement noise, $\vec{\artNoise}_{\time}^{\np}\sim \normal(\vec{0},R)$
	for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:artificial:noise}
        \;
        errors, $\vec{r}_{\time}^{\np}= \vec{\hat{Y}}^{\np}_{\!\time} - \data{\vec{y}}_{\time}$
	for $\np$ in $\seq{1}{\Np}$
        \;
        filter update,
        $\vec{X}_{\time}^{F,\np} = \vec{X}_{\time}^{P,\np} +
              K\big( \vec{r}_{\time}^{\np}+\vec{\artNoise}_{\time}^{\np}\big)$
	for $\np$ in $\seq{1}{\Np}$
        \nllabel{alg:enkf:update}
	\;
	$\loglik_{\time}=\log \big[ \phi\big(\data{\vec{y}}_{\time} \giventh \overline{\vec{Y}}_{\!\time} , \Sigma_{Y} \big) \big]$ where $\phi(\cdot\giventh \vec{\mu},\Sigma)$ is the $\normal(\vec{\mu},\Sigma)$ density.
      }
  \KwOut{
    filter sample, $\vec{X}^{F,1:\Np}_{\time}$, for $n$ in $\seq{1}{N}$;
    log-likelihood estimate, $\loglik^{\mbox{\tiny{EnKF}}}=\sum_{\time=1}^{\Time} \loglik_{\time}$
  }
  \KwCplx{$\bigO(\Np\Unit\Time)$}
\end{algorithm}

@@

\begin{algorithm}[t]
 \caption{\code{bpfilter(P,Np{\argequals}$\Np\!$,block\_list{\argequals}$\blocklist$)}
    using notation from \cref{tab:notation} where \code{P} is a
     `\code{\spatPomp}' object equipped with
     \code{rprocess},
     \code{dunit\_measure},
     \code{rinit},
     \code{obs},
     \code{coef}.
   }\label{alg:bpfilter}
  \KwIn{
    simulator for $f_{\vec{X}_{\time}|\vec{X}_{\time-1}}(\vec{x}_{\time}\given \vec{x}_{\time-1}\giventh\vectheta)$ and $f_{\vec{X}_0}(\vec{x}_{0}\giventh\vectheta)$;
   number of particles, $\Np$;
    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time}\giventh\vectheta)$;
    data, $\data{\vec{y}}_{1:\Time}$;
    parameter, $\vectheta$;
     blocks, $\blocklist_{1:\Block}$;
  }
initialization, $\vec{X}_{0}^{F,\np}\sim {f}_{\vec{X}_{0}}\left(\mydot\giventh{\vectheta}\right)$
for $\np$ in $\seq{1}{\Np}$
\;
\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{ % For each time step
  prediction,
  $\vec{X}_{\time}^{P,\np}\sim {f}_{\vec{X}_{\time}|\vec{X}_{\time-1}}\big(\mydot|\vec{X}_{\time-1}^{F,\np};{\vectheta}\big)$
  for $\np$ in $\seq{1}{\Np}$
    \;
    block weights,
          $\displaystyle \blockweight_{\block,\time}^{\np}=
            \prod_{\unit \in \blocklist_{\block}}
            f_{Y_{\unit,\time}|X_{\unit,\time}}
            \big(
              \data{y}_{\unit,\time}\given X^{P,\np}_{\unit,\time} \giventh \vectheta
            \big)$
    for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
          \nllabel{alg:bpfilter:blockweights}\;
   resampling indices,
      $\resampleIndex^{\np}_{\block,\time}$ with
      $\prob\left[\resampleIndex^{\np}_{\block,\time}=i \right] =
      {\blockweight}^{i}_{\block,\time}
      \Big/\sum_{q=1}^{\Np}\blockweight^{q}_{\block,\time}$
      for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
    \;
    resample,
      $X_{\blocklist_{\block},\time}^{F,\np}=
      X_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}}$
      for $\np$ in $\seq{1}{\Np}$, $\block$ in $\seq{1}{\Block}$
      \nllabel{alg:bpfilter:resample}
    \;
}%End timestep for Loop
\KwOut{
    log-likelihood,
      $\loglik^{\mbox{\tiny{BPF}}}(\vectheta)= \sum_{\time=1}^{\Time}
        \sum_{\block=1}^{\Block}
\log\Big(
        \frac{1}{\Np}
        \sum_{\np=1}^{\Np}
          \blockweight^{\np}_{\block,\time}
      \Big)$, filter particles $\vec{X}^{F,1:\Np}_{1:\Time}$
}
\KwCplx{$\bigO(\Np \Unit \Time)$}
\end{algorithm}

@@

The adapted bagged filter \citep{ionides23} combines many independent particle filters.
This is called \emph{bagging}, (\emph{b}ootstrap \emph{agg}regat\emph{ing}), since a basic particle filter is also called a bootstrap filter.
The adapted distribution is the conditional distribution of the latent process given its current value and the subsequent observation \citep{johansen08}.
In the adapted bagged filter, each bootstrap replicate makes a Monte Carlo approximation to a draw from the adapted distribution.
Thus, in the pseudocode of \cref{alg:abf}, $\vec{X}^{\IF,\rep}_{0:\Time}$ is a Monte Carlo sample targeting the adapted sampling distribution,
\begin{equation}
  \label{adapteddistribution}
f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)\prod_{\time=1}^\Time
f_{\vec{X}_{\time}|\vec{Y}_{\time},\vec{X}_{\time-1}}
(\vec{x}_{\time}\given \data{\vec{y}}_{\time},\vec{x}_{\time-1} \param\vectheta).
\end{equation}
Each adapted simulation replicate is constructed by importance sampling using proposal particles $\{\vec{X}^{\IP,\rep,\np}_{\time}\}$.
The ensemble of adapted simulation replicates are then weighted using data in a spatiotemporal neighborhood of each observation to obtain a locally combined Monte Carlo sample targeting the filter distribution, with some approximation error due to the finite spatiotemporal neighborhood used.
This local aggregation of the bootstrap replicates also provides an evaluation of the likelihood function.

@@

ABF can be combined with the guided intermediate resampling technique used by GIRF to give an algorithm called ABF-IR \citep{ionides23} implemented as \code{abfir}.

@@

  parameter log-likelihoods, 

@@

The \code{measles()} and \code{he10()} models are similar, but differ in details.
For \code{measles()}, the data consist of biweekly counts and parameters are all shared between units, matching the analysis of \citet{park20} and \citet{ionides23}.
For \code{he10()}, data are weekly and parameters can be shared or unit-specific, matching the analysis of \citet{he10} and \citet{ionides22}.
We write the model for the general case where all parameters are unit-specific, noting that it is a relevant data analysis question to determine when parameter dependence on $\unit$ can be omitted.
